---
title: "Fires Data Exploration"
format: html
editor: visual
---

```{r, echo = FALSE, message = FALSE}
library(DBI)
library(dplyr)
library(ggplot2)
library(tigris)
```

## Load Data

```{r}
# Read data
wildfireDataLoc = "C:/Users/kwwur/Documents/personal_projects/wildfires_project/data/FPA_FOD_20170508.sqlite"

con <- dbConnect(RSQLite::SQLite(), wildfireDataLoc)
dbListTables(con)

# dates are not being read in nicely, so pull those in seperately as datetime
wildfires = dbGetQuery(con, 'SELECT * FROM Fires')
wildfire_dates = dbGetQuery(con, 'SELECT datetime(DISCOVERY_DATE) AS DISCOVERY_DATE, datetime(CONT_DATE) AS CONT_DATE, OBJECTID FROM FIRES')

wildfires = left_join(wildfires %>% select(-c(DISCOVERY_DATE, CONT_DATE)), wildfire_dates, by = "OBJECTID")

# create date, year, and month column
date_processing = function(data, col, out_name){
  dates = as.Date(data[[col]])
  years = lubridate::year(dates)
  months = format(dates, "%m-%Y")
  column_names = paste0(out_name, c("_date", "_year", "months"))
  out = data.frame(dates, years, months)
  names(out) = column_names
  return(out)
}

wildfires = cbind(wildfires, date_processing(wildfires, col = "DISCOVERY_DATE", "discovery_processed"))
wildfires = cbind(wildfires, date_processing(wildfires, col = "CONT_DATE", "contained_processed"))


# View data
head(wildfires)

# summary doesn't recognize blob type column, so remove the Shape column
summary(wildfires %>% select(-Shape))
```

### Takeaways from summary

Some columns should be converted to factors:
- STAT_CAUSE_CODE

Some columns should be converted to dates:
- DISCOVERY_DATE
- CONT_DATE


### Address TODOs generated from summary step
```{r}
# convert STAT_CAUSE_CODE to a factor so it is not read as a numeric
wildfires <- wildfires %>% mutate(STAT_CAUSE_CODE = as.factor(STAT_CAUSE_CODE))

# this date conversion removes the time from the string, but most time information is listed as 00:0000 so that wasn't going to be useful information
wildfires <- wildfires %>% mutate(DISCOVERY_DATE = as.Date(DISCOVERY_DATE), 
                                  CONT_DATE = as.Date(CONT_DATE))

```


## Missing Data

### Check for Missing Data

```{r}
missing_count = colSums(is.na(wildfires))

data.frame(Percent_Missing = missing_count/nrow(wildfires)*100) %>%
  mutate(Percent_Missing = paste0(round(Percent_Missing, 2), "%"))

cols_w_missing = names(wildfires)[missing_count != 0]
```

### Takeaways from missing data

Some columns have too much missingness to ignore or reliably impute. Those are:

- LOCAL_FIRE_REPORT_ID
- LOCAL_INCIDENT_ID
- FIRE_CODE
- FIRE_NAME
- ICS_209_INCIDENT_NUMBER
- ICS_209_NAME
- MTBS_ID
- MTBS_FIRE_NAME
- COMPLEX_NAME
- DISCOVERY_TIME

COUNTY is the less nicely formatted version of FIPS_CODE, so we can remove the COUNTY column without losing any information. 

The CONT... columns have a large number of NA values, but NA values may indicate that they are not multi-day fires. The reason for these NA's is not made clear in the data source site. I will keep these columns, but limit to fires with complete informtion when making use of these columns.   
- CONT_DATE
- CONT_DOY
- CONT_TIME

COUNTY, FIPS_CODE, and FIPS_NAME are missing in over one third of entries, but they may be determined using lat/lon coordinates. 
- COUNTY (but we're going to remove this anyway)
- FIPS_CODE
- FIPS_NAME

### Address TODOs from missing data
```{r}
# remove columns we will not be able to use because there is too much missing information
wildfires <- wildfires %>% select(-c(LOCAL_FIRE_REPORT_ID, 
                                     LOCAL_INCIDENT_ID, 
                                     FIRE_CODE, 
                                     FIRE_NAME, 
                                     ICS_209_INCIDENT_NUMBER, 
                                     ICS_209_NAME,
                                     MTBS_ID, 
                                     MTBS_FIRE_NAME, 
                                     COMPLEX_NAME, 
                                     DISCOVERY_TIME)
                                  )

# remove COUNTY column since we can get that information from FIPS_CODE
wildfires <- wildfires %>% select(-COUNTY)

# get US county geometries
US_counties <- tigris::counties()

# select lat/lon coordinates and format as an sf points object
lat_lon_coords = wildfires %>% select(LATITUDE, LONGITUDE)
latlon_pts <- sf::st_as_sf(lat_lon_coords, coords = c("LONGITUDE", "LATITUDE"), crs = 4326)

# transform counties and points geometries to have the same coordinate reference system
US_counties <- sf::st_transform(US_counties, crs = sf::st_crs(latlon_pts))

# find the index of the county each point falls in 
intersects <- sf::st_intersects(latlon_pts, US_counties)

# there were two points that were matched to two counties. In those cases, I just selected the first one. 
county_indexes <- lapply(intersects, function(x){x[1]}) %>% unlist()

# identify the name and fips code of the matched county
latlon_counties <- US_counties %>% 
  select("FIPS_NAME_calculated" = "NAME", 
         "FIPS_CODE_calculated" = "COUNTYFP") %>%
  as.data.frame() %>%
  select(-geometry)

point_counties <- latlon_counties[county_indexes,]

# join information back with the wildfire dataset and only use calculated values if original values are missing
wildfires <- cbind(wildfires, point_counties)
wildfires <- wildfires %>% mutate(FIPS_NAME = ifelse(is.na(FIPS_NAME), FIPS_NAME_calculated, FIPS_NAME),
                                  FIPS_CODE = ifelse(is.na(FIPS_CODE), FIPS_CODE_calculated, FIPS_CODE)) %>%
  select(-c(FIPS_NAME_calculated, FIPS_CODE_calculated))
row.names(wildfires) <- NULL

# add lat/lon points since that could be useful
wildfires <- cbind(wildfires, latlon_pts %>% rename("latlon_geometry" = "geometry"))

# add county geometry since that could be useful
# TODO: this needs to be done in SQL
# US_counties <- US_counties %>% select(-NAME) %>% rename("county_geometry" = "geometry")
# which(US_counties$COUNTYFP == wildfires$FIPS_CODE)
# 
# wildfires <- left_join(wildfires, US_counties %>% select(-NAME) %>% rename("county_geometry" = "geometry"), by = c("FIPS_CODE" = "COUNTYFP"))

```

Now we want to double check that the county information is now complete

```{r}
# Confirm that FIPS_NAME and FIPS_CODE are now complete
missing_count = colSums(is.na(wildfires %>% select(FIPS_NAME, FIPS_CODE)))
missing_count

```

## Histograms

```{r}

ggplot_hist = function(dataset = wildfires,
                       column_nm,
                       binwidth = NULL) {
  ggplot(dataset) +
    geom_histogram(
      aes(get(column_nm)),
      binwidth = binwidth,
      alpha = .5,
      color = "black"
    ) + 
    xlab(column_nm)
  
}

ggplot_hist(column_nm = "FIRE_YEAR", binwidth = 1)
ggplot_hist(column_nm = "FIRE_SIZE")
ggplot_hist(data = wildfires %>% select(-Shape) %>% mutate(log_fire_size = log(FIRE_SIZE)), column_nm = "log_fire_size")
ggplot_hist(column_nm = "discovery_processed_date")
ggplot_hist(column_nm = "contained_processed_date")

```

### Takeaways

- Relatively consistent number of fires per year
- Seems to be a large number of smaller fires with a few VERY large fires. TODO: sniff check on the largest fires 


## Summary Tables

```{r}
# confirm OBJECTID and FODID are unique
length(unique(wildfires$OBJECTID)) == nrow(wildfires)
length(unique(wildfires$FOD_ID))

# this id is not unique
length(unique(wildfires$FPA_ID))

# create table of these columns
length(unique(wildfires$SOURCE_SYSTEM_TYPE))
length(unique(wildfires$SOURCE_SYSTEM))
length(unique(wildfires$NWCG_REPORTING_AGENCY))
length(unique(wildfires$STAT_CAUSE_CODE))
length(unique(wildfires$STAT_CAUSE_DESCR))
length(unique(wildfires$FIRE_SIZE_CLASS))
length(unique(wildfires$OWNER_CODE))
length(unique(wildfires$OWNER_DESCR))
length(unique(wildfires$STATE))

# there are too many of these rows to create tables of
length(unique(wildfires$NWCG_REPORTING_UNIT_ID))
length(unique(wildfires$NWCG_REPORTING_UNIT_NAME))
length(unique(wildfires$SOURCE_REPORTING_UNIT))
length(unique(wildfires$SOURCE_REPORTING_UNIT_NAME))
length(unique(wildfires$LOCAL_FIRE_REPORT_ID))
length(unique(wildfires$LOCAL_INCIDENT_ID))
length(unique(wildfires$FIRE_CODE))
length(unique(wildfires$FIRE_NAME))
length(unique(wildfires$ICS_209_INCIDENT_NUMBER))
length(unique(wildfires$ICS_209_NAME))
length(unique(wildfires$MTBS_ID))
length(unique(wildfires$MTBS_FIRE_NAME))
length(unique(wildfires$COMPLEX_NAME))
length(unique(wildfires$DISCOVERY_DATE))
length(unique(wildfires$DISCOVERY_DOY))
length(unique(wildfires$DISCOVERY_TIME))
length(unique(wildfires$CONT_DATE))
length(unique(wildfires$CONT_DOY))
length(unique(wildfires$CONT_TIME))
length(unique(wildfires$FIRE_SIZE))
length(unique(wildfires$COUNTY))
length(unique(wildfires$FIPS_CODE))
length(unique(wildfires$FIPS_NAME))
```
```{r}
create_summary_table = function(data, col_nm){
  out = data %>% group_by(get(col_nm)) %>% summarize("number of observations" = n())
  return(out)
}

create_summary_table(data = wildfires, col_nm = "SOURCE_SYSTEM_TYPE")
create_summary_table(data = wildfires, col_nm = "SOURCE_SYSTEM")
create_summary_table(data = wildfires, col_nm = "NWCG_REPORTING_AGENCY")
create_summary_table(data = wildfires, col_nm = "STAT_CAUSE_CODE")
create_summary_table(data = wildfires, col_nm = "STAT_CAUSE_DESCR")
create_summary_table(data = wildfires, col_nm = "FIRE_SIZE_CLASS")
create_summary_table(data = wildfires, col_nm = "OWNER_CODE")
create_summary_table(data = wildfires, col_nm = "OWNER_DESCR")
create_summary_table(data = wildfires, col_nm = "STATE")
```



